Bayesian statistics for repeated measures

### 2020-10-21 Manuscripts/Manuscript_AM_v1.Rmd
Updated

The study of the temporal changes in a variable of interest (a longitudinal analysis) presents advantages over a cross-sectional study as it reduces the number of subjects required to achieve a certain statistical power[@guo2013], it enables a time-correlated view of the phenomenon of interest, and allows to explore multiple relationships between different variables. Typically, measurements are taken across multiple timepoints on the same subjects within  different group(s). Examples of this type of approach in biomedical research include clinical studies on breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].

Researchers typically employ a $frequentist$ approach to analyze longitudinal data. Such type of analysis is a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). Although the usual approach, rm-ANOVA assumes that the data fulfills certain requisites: Constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004]. Moreover, this approach restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1. with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009]. Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a $frequentist$ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself. c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies. With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.

>>>>

A longitudinal study is defined as that were a variable of interest is measured repeatedly in a group (or groups) of subjects. In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment (or treatments) across time, rather than analyzing the information at a single timepoint (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015], where in the first case  weekly measurements of skin toxicities in breast cancer patients with radiation-induced dermatitis were taken for up to 8 weeks, and mouth opening in head and neck cancer patients was measured at 6,12, 18, 24 and 36 months after radiotherapy (RT) in the latter. Similar examples are found in studies of tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides the previously mentioned time-effect evolution, it allows to determine the intra-variability of the response between different subjects [@guo2013; @fitzmaurice2012].

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data. Such type of analysis is based on a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). However,a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010]. Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010]. Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].

However, both LMEMs and rm-ANOVA restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1. with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009; @roblyer2011,@]. Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a _frequentist_ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself. c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies. With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
